---
title: "Munich in numbers"
author: "CU"
date: "10/6/2019"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 2
---
```{r setup, include=FALSE}
#   github_document
   # toc: true
    # toc_depth: 2

library(knitr)         # markdown enhancements
library(kableExtra)    # nice rmarkdown tables
library(RCurl)         # download data
library(readxl)        # read Excel
library(tidyverse)     # tidyverse universe (dplyr, ggplot ...)


knitr::opts_chunk$set(echo = TRUE)
##########################
# preparations
rm(list=ls())
gc()

# url
url = 'http://www.mstatistik-muenchen.de/monatszahlenmonitoring/export/xlsx/mzm_export_alle_monatszahlen.xlsx'
destExcelFile = "file.xlsx"

download.file(url = url,
              destfile=destExcelFile,
              method="libcurl",
              mode = "wb")

path = destExcelFile
dat <- path %>% 
  excel_sheets() %>% 
  set_names() %>% 
  map(read_excel, path = path)

rm(path, destExcelFile, url)

##########################
```

# Brief exploratory analysis

This is a brief EDA of Munich stats.

## Data 

### source
Thanks to **Statistische Amt München** for making the data available on a monthly basis. They make way more data available, these are solely main KPIs. Source of data: <http://www.mstatistik-muenchen.de/monatszahlenmonitoring/export/export.php>. Official Data exploration portal: <http://www.mstatistik-muenchen.de/monatszahlenmonitoring/atlas.html?indicator=i158&date=Jan&select=20,19&select2=JAHR&indicator2=i0>. Data here is extracted from Excel, processed into data.frames and correlation charts


### raw
The raw data is a list of data-frames

```{r raw_data_overview, message=FALSE, collapse=FALSE}
as.data.frame(lapply(dat, dim)) %>% 
  rownames_to_column() %>%
  kable()
```

### aggregated
We'll now compress the **original disparated data** in one compact data.frame
For that, we create dynamic column names and aggregate the monthly values
The final result will be a large matrix (~300 columns) where each column represents a variable and the rows are the monthly values (20 yrs x 12 months)


```{r prepare_large_aggregation, warning=FALSE}
##########################
# looping through all the Excel sheets
dsFinal <- data.frame()
for (i in 1:length(names(dat))) {
  SECTIONNAME <- names(dat)[i]
  dtmp <- (dat[SECTIONNAME])[[1]]
  # if the current section has values in it
  if ('WERT' %in% names(dtmp)) {
    # creating dynamic column names and assign monthly values as cols
    dtmp_mo <- dtmp %>%
      select(MONATSZAHL, AUSPRAEGUNG, MONAT, WERT) %>%
      mutate(colname=paste(SECTIONNAME, MONATSZAHL, AUSPRAEGUNG, sep = " - ")) %>%
      select(MONAT, colname, WERT) %>%
      spread(MONAT, WERT) 
    rownames(dtmp_mo) <- dtmp_mo$colname
    dtmp_mo$colname <- NULL
    # transpose monthly values to rows
    dtmp_mo <- t(dtmp_mo)
    # back from tibble to dataframe
    dtmp_mo <-as.data.frame(dtmp_mo)
    dtmp_mo$month <- rownames(dtmp_mo)
    
    if (nrow(dsFinal)==0){
      # if the final construct doesn't exist, simply create it
      dsFinal <- dtmp_mo
    } else {
      # if we have values in it already, merge them (full join)  using the month column
      dsFinal <- dtmp_mo %>%
        full_join(dsFinal, by = "month")
    }
  }
}
# cleanup
rm(dtmp, dtmp_mo, i, SECTIONNAME)

rownames(dsFinal) <- dsFinal$month
dsFinal$month <- NULL
dim(dsFinal)
# dsFinal[1:10,1:2] %>% kable()
# colMeans(dsFinal, na.rm = T) %>% kable()
# str(dsFinal)
##########################
```

# General Trends (monthly)
## Cinema visitors

```{r charts_kinos, message=FALSE}
##########################
# exploratory charts
dat$KINOS %>%
  mutate(Month = as.Date(paste0(MONAT,'01'), format = '%Y%m%d')) %>%
  ggplot(., aes(x=Month, y = WERT, group=AUSPRAEGUNG)) + 
    geom_line() + 
    facet_wrap(~AUSPRAEGUNG) +
    theme_classic()+ 
    geom_smooth() + scale_y_continuous(labels = scales::comma) + ggtitle("Cinema Visitors")
```

## weather trends 
### Sun & temperature

```{r charts_weather, message=FALSE}
dat$WITTERUNG %>%
  filter(MONATSZAHL %in% c("Sonnenschein", "Lufttemperatur")) %>%
  mutate(Month = as.Date(paste0(MONAT,'01'), format = '%Y%m%d')) %>%
  ggplot(., aes(x=Month, y = WERT, group=AUSPRAEGUNG)) + 
  geom_line() + 
  facet_wrap(~AUSPRAEGUNG, scales = "free") +
  theme_classic()+ 
  geom_smooth() + scale_y_continuous(labels = scales::comma) + ggtitle("monthly weather trends")
```

### Rain

```{r charts_weather_rain, message=FALSE}
dat$WITTERUNG %>%
  filter(MONATSZAHL %in% c("Niederschlag")) %>%
  mutate(Month = as.Date(paste0(MONAT,'01'), format = '%Y%m%d')) %>%
  ggplot(., aes(x=Month, y = WERT, group=AUSPRAEGUNG)) + 
  geom_line() + 
  facet_wrap(~AUSPRAEGUNG, scales = "free", nrow=2) +
  theme_classic()+ 
  geom_smooth() + scale_y_continuous(labels = scales::comma) + ggtitle("monthly weather trends")
```

## Tourists

```{r charts_tourism, message=FALSE}
dat$TOURISMUS %>%
  filter(MONATSZAHL %in% c("Gäste")) %>%
  mutate(Month = as.Date(paste0(MONAT,'01'), format = '%Y%m%d')) %>%
  ggplot(., aes(x=Month, y = WERT, group=AUSPRAEGUNG)) + 
  geom_line() + 
  facet_wrap(~AUSPRAEGUNG, scales = "free", nrow=3) +
  theme_classic()+ 
  geom_smooth() + scale_y_continuous(position = "right", labels = scales::comma) + ggtitle("monthly trend - Turist guests")
```

## Population

### German population 2010-today

```{r charts_population_germans_only, message=FALSE}
dsFinal %>%
  rownames_to_column() %>%
  select(rowname, `german population` =  `BEVÖLKERUNG - Geschlecht und Staatsangehörigkeit - Deutsche insgesamt`) %>%
  mutate(Month = as.Date(paste0(rowname,'01'), format = '%Y%m%d')) %>%
  filter(Month>='2010-01-01') %>%
  ggplot(aes(x = Month))  + 
    geom_line(aes(y = `german population`)) + 
    theme_minimal() + 
    scale_x_date(date_breaks = "1 year", date_labels = "%Y") + 
    scale_y_continuous(position = "right", labels = scales::comma) 
```

### Foreigners 2010-today

```{r charts_population_foreigners_only, message=FALSE}
dsFinal %>%
  rownames_to_column() %>%
  select(rowname, `foreigners` = `BEVÖLKERUNG - Geschlecht und Staatsangehörigkeit - Ausländer/innen insgesamt`) %>%
  mutate(Month = as.Date(paste0(rowname,'01'), format = '%Y%m%d')) %>%
  filter(Month>='2010-01-01') %>%
  ggplot(aes(x = Month))  + 
    geom_line(aes(y = foreigners), linetype="twodash") + 
    theme_minimal() + 
    scale_x_date(date_breaks = "1 year", date_labels = "%Y") + 
    scale_y_continuous(position = "right", labels = scales::comma) 
```

### Inhabitants (all & gender)

```{r charts_population_all, message=FALSE}
dat$BEVÖLKERUNG %>%
  filter(MONATSZAHL %in% c("Geschlecht und Staatsangehörigkeit")) %>%
  filter(str_detect(AUSPRAEGUNG, 'Einwohner')) %>%
  mutate(Month = as.Date(paste0(MONAT,'01'), format = '%Y%m%d')) %>%
  arrange(-WERT) %>% # sort
  mutate_at(vars(AUSPRAEGUNG), funs(factor(., levels=unique(.)))) %>% # convert to factor
  ggplot(., aes(x=Month, y = WERT, group=AUSPRAEGUNG)) + 
  geom_line() + 
  facet_wrap(~AUSPRAEGUNG) +
  theme_classic()+ 
  scale_y_continuous(labels = scales::comma) + ggtitle("Munich Inhabitants (male, female, all)")
```

### German inhabitants (all & gender)

```{r charts_population_german, message=FALSE}
dat$BEVÖLKERUNG %>%
  filter(MONATSZAHL %in% c("Geschlecht und Staatsangehörigkeit")) %>%
  filter(str_detect(AUSPRAEGUNG, 'Deutsche')) %>%
  mutate(Month = as.Date(paste0(MONAT,'01'), format = '%Y%m%d')) %>%
  arrange(-WERT) %>% # sort
  mutate_at(vars(AUSPRAEGUNG), funs(factor(., levels=unique(.)))) %>% # convert to factor
  ggplot(., aes(x=Month, y = WERT, group=AUSPRAEGUNG)) + 
  geom_line() + 
  facet_wrap(~AUSPRAEGUNG) +
  theme_classic()+ 
  scale_y_continuous(labels = scales::comma) + ggtitle("Germans in Munich (male, female, all)")
```

### Foreign inhabitants (all & gender)

```{r charts_population_foreign, message=FALSE}
dat$BEVÖLKERUNG %>%
  filter(MONATSZAHL %in% c("Geschlecht und Staatsangehörigkeit")) %>%
  filter(str_detect(AUSPRAEGUNG, 'Ausländer')) %>%
  mutate(Month = as.Date(paste0(MONAT,'01'), format = '%Y%m%d')) %>%
  arrange(-WERT) %>% # sort
  mutate_at(vars(AUSPRAEGUNG), funs(factor(., levels=unique(.)))) %>% # convert to factor
  ggplot(., aes(x=Month, y = WERT, group=AUSPRAEGUNG)) + 
  geom_line() + 
  facet_wrap(~AUSPRAEGUNG) +
  theme_classic()+ 
  scale_y_continuous(labels = scales::comma) + ggtitle("Foreigners in Munich (male, female, all)")
```


### Female share
#### among german nationals

```{r female_share_germans, message=FALSE}

dsFinal %>% 
  rownames_to_column() %>%
  mutate(
    year = substring(rowname, 0, 4), 
    'german_female' = `BEVÖLKERUNG - Geschlecht und Staatsangehörigkeit - Deutsche weiblich`,
   'german_all' = dsFinal$`BEVÖLKERUNG - Geschlecht und Staatsangehörigkeit - Deutsche insgesamt`) %>%
  select(year, german_female, german_all) %>%
  group_by(year) %>%
  summarize(german_all=round(mean(german_all, na.rm = TRUE),0),
            german_female = round(mean(german_female, na.rm = TRUE),0)
            ) %>%
  mutate(female_share = round(german_female/german_all, 2)) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))


```

#### among foreigners

```{r female_share_foreigners, message=FALSE}

plot(
    rownames(dsFinal) 
    , dsFinal$`BEVÖLKERUNG - Geschlecht und Staatsangehörigkeit - Ausländer weiblich`  / dsFinal$`BEVÖLKERUNG - Geschlecht und Staatsangehörigkeit - Ausländer/innen insgesamt`
    , xlab = "month"
    , ylab = "% Female"
     , main="% Female among Foreigners", 
    type="o")


```


### Family status

```{r charts_population, message=FALSE}
dat$BEVÖLKERUNG %>%
  filter(MONATSZAHL %in% c("Familienstand")) %>%
  mutate(Month = as.Date(paste0(MONAT,'01'), format = '%Y%m%d')) %>%
  arrange(-WERT) %>% # sort
  mutate_at(vars(AUSPRAEGUNG), funs(factor(., levels=unique(.)))) %>% # convert to factor
  ggplot(., aes(x=Month, y = WERT, group=AUSPRAEGUNG)) + 
  geom_line() + 
  facet_wrap(~AUSPRAEGUNG, scales = "free") +
  theme_classic()+ 
  scale_y_continuous(labels = scales::comma) + ggtitle("monthly trend - population by family status")
#ggsave("monthly trend - population by family status.png", dpi=400, dev='png', height=4, width=5, units="in", scale = 2)
```

### Age groups

```{r charts_population_by_age, message=FALSE}
dat$BEVÖLKERUNG %>%
    filter(MONATSZAHL %in% c("Altersgruppen")) %>%
    mutate(Month = as.Date(paste0(MONAT,'01'), format = '%Y%m%d')) %>%
    arrange(-WERT) %>% # sort
    mutate_at(vars(AUSPRAEGUNG), funs(factor(., levels=unique(.)))) %>% # convert to factor
    ggplot(., aes(x=Month, y = WERT, group=AUSPRAEGUNG)) + 
    geom_line() + 
    facet_wrap(~AUSPRAEGUNG, scales = "free") +
    theme_classic()+ 
    geom_smooth() + scale_y_continuous(labels = scales::comma) + ggtitle("Population by age group")
```

## Religion

```{r charts_population_religion, message=FALSE}
dat$BEVÖLKERUNG %>%
    filter(MONATSZAHL %in% c("Religionszugehörigkeit")) %>%
    mutate(Month = as.Date(paste0(MONAT,'01'), format = '%Y%m%d')) %>%
    ggplot(., aes(x=Month, y = WERT, group=AUSPRAEGUNG)) + 
    geom_line() + 
    facet_wrap(~AUSPRAEGUNG, scales = "free") +
    theme_classic()+ 
    geom_smooth() + scale_y_continuous(labels = scales::comma) + ggtitle("Religion in Munich")
```


## Unemployment (gender & nationality)
### absolute numbers

```{r charts_unemployed_absolute, message=FALSE}
dat$ARBEITSMARKT %>%
  filter(MONATSZAHL %in% c("Arbeitslose")) %>%
  filter(AUSPRAEGUNG %in% c("insgesamt","Frauen","Männer","Deutsche","Deutsche", "Ausländer/innen", "Langzeitarbeitslose")) %>%
  mutate(Month = as.Date(paste0(MONAT,'01'), format = '%Y%m%d')) %>%
  arrange(-WERT) %>% # sort
  mutate_at(vars(AUSPRAEGUNG), funs(factor(., levels=unique(.)))) %>% # convert to factor
  ggplot(., aes(x=Month, y = WERT, group=AUSPRAEGUNG)) + 
  geom_line() + 
  facet_wrap(~AUSPRAEGUNG, scales = "free") +
  theme_classic()+ 
  scale_y_continuous(labels = scales::comma) + ggtitle("Unemployed in Munich (#)")
```

### as percentages 

```{r charts_unemployed_percent, message=FALSE}
dat$ARBEITSMARKT %>%
  filter(MONATSZAHL %in% c("Arbeitslosenquote")) %>%
  filter(AUSPRAEGUNG %in% c("abh. ziv. Erwerbspersonen", "alle ziv. Erwerbspersonen")) %>%
  mutate(Month = as.Date(paste0(MONAT,'01'), format = '%Y%m%d')) %>%
  arrange(-WERT) %>% # sort
  mutate_at(vars(AUSPRAEGUNG), funs(factor(., levels=unique(.)))) %>% # convert to factor
  ggplot(., aes(x=Month, y = WERT, group=AUSPRAEGUNG)) + 
  geom_line() + 
  facet_wrap(~AUSPRAEGUNG, scales = "free") +
  theme_classic()+ 
  scale_y_continuous(labels = scales::comma) + ggtitle("Unemployment rate in Munich")
```

# Counting NAs (missing data)

```{r prepare_count_of_NAs}

##########################
# count NAs
# absolute number of missing values
nasCount <- dsFinal %>%
  summarise_all(funs(sum(is.na(.))))
# relative count
nasCount <- nasCount/nrow(dsFinal)
# transpose percent of missing values
nasCount <- as.data.frame(t(nasCount))
nasCount$field <- rownames(nasCount)
names(nasCount) <- c('NAs_proc', 'field_name')
# done
head(nasCount) %>% kable()
nasFinal <- data.frame("field_name" =  nasCount$field_name, "NAs_proc" = nasCount$NAs_proc)
plot(nasFinal$NAs_proc, main="Percent of missing values for all variables")
abline(h=0.1, col="blue")
rm(nasCount)
##########################
```
we will **cut all variables filled below 10%** (so all above the blue line).


# Explore correlations
## simple use of cor()

```{r prepare_correlation_cor}
##########################
# explore correlation
dsCor <- cor(dsFinal, use = "pairwise.complete.obs")

dsCor[1:5, 1:5]

# explore correlations with R https://drsimonj.svbtle.com/exploring-correlations-in-r-with-corrr
diag(dsCor) <- NA
col_has_over_90 <- apply(dsCor, 2, function(x) any(x > .9))

length(col_has_over_90)

#( dsCor[,col_has_over_90] )

```

## the correlate package

```{r prepare_correlation_corrr, warning=FALSE}

library(corrr)
correlate(dsFinal, use = "pairwise.complete.obs")

# dsFinal %>% 
#   correlate(use = "pairwise.complete.obs") %>% 
#   focus(3:5, mirror = TRUE) %>% 
#   network_plot()

# above 98% pearson correlation
dsFinal %>% 
  correlate(use = "pairwise.complete.obs") %>% 
  stretch() %>% 
  filter(abs(r)>0.98) %>% 
  head(20)

# how many does each correlate with
dsFinal %>% 
  correlate(use = "pairwise.complete.obs") %>% 
  stretch() %>% 
  filter(abs(r)>0.98) %>%
  group_by(x) %>%
  count() %>% 
  arrange(-n) %>% 
  head(20)

# split the correlated variables by section/subsections
dsFinal %>% 
  correlate(use = "pairwise.complete.obs") %>% 
  stretch() %>% 
  filter(abs(r)>0.98) %>%
  mutate(x_field = x, y_field = y) %>%
  separate(x, into = c("x_sec", "x_subsection", "x_subsubsection"), sep = " - ") %>%
  separate(y, into = c("y_sec", "y_subsection", "y_subsubsection"), sep = " - ") %>%
  filter(x_sec!=y_sec)-> dsAgg

# main correlated sections
( table(dsAgg$x_sec, dsAgg$y_sec) )

# dsFinal %>% 
#   correlate(use = "pairwise.complete.obs") %>% 
#   network_plot(min_cor = 0.98)

  

dsAgg %>%
  inner_join(nasFinal, by = c("x_field"="field_name")) %>%
  filter(NAs_proc <0.1) %>%
  select(x_field, y_field, r, NAs_proc) -> dsAggFiltered
(dsAggFiltered)
##########################
```

# Next steps

- we look at cross-topic correlations (assuming that inter-topic correlations are logical: highest to lowest temperature f.e.)
- explore some highly correlated numbers
- **BEWARE OF CONFOUNDING VARIABLES**

## Highly Correlated variables
These are pure sample examples out of  ~100

```{r prepare_correlation_charts, warning=FALSE}
plot(dsFinal$`BEVÖLKERUNG - Altersgruppen - Rentner/innen (65 J. und älter)`, dsFinal$`SOZIALE LEISTUNGEN - Empfänger nach SGB XII - Empfänger/innen insges.`,
     main="Old-age pensioner and social welfare recipients")
plot(dsFinal$`BEVÖLKERUNG - Familienstand - geschieden`, dsFinal$`KFZ-Bestand - Pkw-Kraftstoffarten - Diesel gesamt`,
     main="Divorces and Diesel cars")
```

## linear models
Handworkers by number of diesel cars and total population

```{r prepare_linear_models, warning=FALSE}
model.lm <- lm(`WIRTSCHAFT - Handwerksbetriebe - Handwerk f. den priv. Bedarf` ~ `KFZ-Bestand - Pkw-Kraftstoffarten - Diesel gesamt` + `BEVÖLKERUNG - Geschlecht und Staatsangehörigkeit - Einwohner/innen insgesamt`, data = dsFinal)
summary(model.lm)
plot(model.lm)
```


:)
